---
title: "question06"
author: "Kyle Alvarez"
format: pdf
---

# Example Code To Be Reduced
```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

# Example Code Reduced to a Function
> Q6. How would you generalize the code above to work with any set of input protein structures

```{r}
analyzePDI <- function(kinase){ # Takes a kinase as its argument
  structure <- read.pdb(kinase) # Reads the protein data bank of the kinase and stores it within the variable structure
  structure.chainA <- trim.pdb(structure, chain="A", elety="CA") # Trims the PDB object of the kinase to get its structure at the specified chain and elety
  structure.b <-  structure.chainA$atom$b # Assigns the specific atom from the proteins's Chain A to a different variable
  plotb3(structure.b, sse=structure.chainA, type="l", ylab="Bfactor") # Plots the data of Residue vs Bfactor from the structure.b variable in a line graph
}

```
In order to use this function all you have to do is type `analyzePDI()` and insert the kinase/protein that you want to be analyzed in quotations within the parentheses of the argument.

The output of the function is a line graph that plots the residue against the Bfactor of a kinase/protein.



# Using the simplified function for the example kinases
Tests the function against the example kinases to see if the graphs are the same, in this case it is. Hence we have a simplified function in which we can use to plot different kinases instead of having to manually type out every step subsequently.
```{r}
analyzePDI("4AKE")
analyzePDI("1AKE")
analyzePDI("1E4Y")
```

